---
title: "Dry Flies Complete Analysis"
author: "Isabeau Lewis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Question 1 Analysis: Investigating differential responses to dessication in four populations of *Drosophila mojavensis* by analyzing changes in cuticular hydrocarbon (CHC) composition.

## Part 1: PCA

We set out to determine whether **population origin**, **host plant**, or **desiccation exposure time**, will result in differential stress responses in *D. mojavensis*. We will analyze our data by performing a principal component analysis and then plotting the most influential components, relative to the three mentioned variables. This will allow us to conclude whether the selected variables yield differential responses and which variables displayed more significant separation of the CHC content values.

Before inputting the data, a unix script was performed to change the csv. The original dataset has 'F' indicating a female *D. mojavensis*. When inputting the dataset into R, these were switched to 'FALSE.' Thus, we used unix and regular expressions to alter this dataset before it was inputted:

$ cd Users/yourname/path/to/csv/file
$ sed 's/^\([^,]*,[^,]*,\)F\([^,]*,\)/\1female\2/' chcData.csv > chcData1.csv

This unix script changes the csv (original: "chcData.csv") and saves it under a new filed named "chcData1.csv." All subsequent analyses will be done using the edited dataframe.

# Formatting Dataset

Inputting dataset and importing necessary packages:

```{r}
library(ggplot2)
library(gridExtra)
library(dplyr)
library(cowplot)

theme_set(theme_classic())

CHC <- read.csv("./data/CHC_dat.csv")
```

 
Checking data for any errors:

```{r}
dim(CHC)
```

```{r}
head(CHC)
```

```{r}
tail(CHC)
```

```{r}
summary(CHC)
```

```{r}
str(CHC)
```

There were some issues with inputting the raw data into R. Major points of concern:
- Row 117 does not have a sample name and many NAs, so going to drop that row
- Extra rows were added into the edges of the data frame

```{r}
CHC1 <- CHC %>%
  dplyr::select(-c(starts_with("X"), pop, sex, cactus, exposure, SAMPLE)) %>% #Removing additional columns, deleting unecessary categorical data
  na.omit() %>% #Omitting bottom row with no data
  mutate_at('C.37.br.alkene', as.numeric) %>% #Making column numeric rather than a character
   mutate_all(~ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x)) #Removing any NAs introduced by character -> numeric

```

```{r}
tail(CHC1)
```

```{r}
dim(CHC1)
```

```{r}
sum(is.na(CHC1$C.37.br.alkene)) #Checking to see how many NAs are left in our dataframe:
```

Checking the normality of the dataset:

```{r}
# loop through each column in the dataframe
for (col in 1:ncol(CHC1)) {
  # calculate the kurtosis using the kurtosis function from the moments package
  kurt <- moments::kurtosis(CHC1[, col])
  
  # check if kurtosis falls between 2 and 4
  if (kurt > 1.5 & kurt < 5) {
    print(paste("Column", col, "is normal"))
  } else {
    print(paste("Column", col, "is not normal"))
  }
}
```

```{r}
# loop through each column in the dataframe
for (col in 1:ncol(CHC1)) {
  # calculate the kurtosis 
  skew <- moments::skewness(CHC1[, col])
  
  # check if kurtosis falls between 2 and 4
  if (kurt > 1.5 & kurt < 5) {
    print(paste("Column", col, "is normal"))
  } else {
    print(paste("Column", col, "is not normal"))
  }
}
```

Based on the kurtosis and skewness of each of the columns within our dataframe, columns 5 and 28 are points of concern within our dataframe do not fit a normal distribution as well. This dataset represents changes in expression of a variety of hydrocarbons and thus, biologically, it makes sense for it not to be normally distributed. We scale the data using z-scores to run our PCA and are therefore making the assumption that this will reduce the impacts of any non-normal data.

Correlation table:

```{r}
round(cor(CHC1),3)
```

 
# Principal Component Analysis

Scaling the data using standards (z-scores).

```{r}
CHC1 <- apply(CHC1, 2,  function(x) (x - mean(x)) / sd(x)) # Applying the scaling function to each column in the dataframe
```

```{r}
CHCPCA <- princomp(CHC1,cor = F)
```


Examining the structure of the PCA analysis:

```{r}
str(CHCPCA)
```

```{r}
summary(CHCPCA)
```

Removing NA row from CHC so we can bind the data (same as CHCDataFixing.R code chunk above):

```{r}
CHC <- CHC %>%
  dplyr::select(-c(starts_with("X"))) %>%
  na.omit()
head(CHC)
```

```{r}
CHCDat<-cbind(CHC,CHCPCA$scores)
head(CHCDat)
```


PC Loadings: Eigenvalues (Representation of variation explained by each PC)

```{r}
PCloadings <- data.frame(Component=c(1:30),Eigenvalue=CHCPCA$sdev^2)
ggplot(aes(x=Component,y=Eigenvalue),data=PCloadings) + geom_point() + geom_line()
```

**Fig. 1.** Scree Plot: Eigenvalues (PC loadings) associated with each of the 30 principal components.  


Since PC 1-4 represent the most variation of the 30 components, we will focus on these 4 components in our analysis. 


Renaming the columns for figure formatting:

```{r}
CHCDat <- CHCDat %>% rename(Population = pop) %>% rename(Cactus = cactus) %>% rename(Exposure = exposure)

# Changing exposure time to a factor for colour visualization
CHCDat <- CHCDat %>% mutate(Exposure=as.factor(Exposure))
```


Comparing the bivariate plots relative to **population origin**: 

```{r}
p1_2 <- ggplot(aes(x=Comp.2,y=Comp.1, colour=Population), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 2") + theme(legend.position = "none")

p1_3 <- ggplot(aes(x=Comp.3,y=Comp.1, colour=Population), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 3") + theme(legend.position = "none")

p1_4 <- ggplot(aes(x=Comp.4,y=Comp.1, colour=Population), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 4") + theme(legend.position = "none")

p2_3 <- ggplot(aes(x=Comp.3,y=Comp.2, colour=Population), data=CHCDat) + geom_point() + ylab("PC 2") + xlab("PC 3") + theme(legend.position = "none")

p2_4 <- ggplot(aes(x=Comp.4,y=Comp.2, colour=Population), data=CHCDat) + geom_point() + ylab("PC 2") + xlab("PC 4") + theme(legend.position = "none")

p3_4 <- ggplot(aes(x=Comp.4,y=Comp.3, colour=Population), data=CHCDat) + geom_point() + ylab("PC 3") + xlab("PC 4") + theme(legend.position = "none")

PCPlot_Pop <- grid.arrange(p1_2, p1_3, p1_4, p2_3, p2_4, p3_4, nrow=2)

```

**Fig. 2.** All possible combinations of principle components 1-4 with points coloured by population origin (red = OPNM, green = PO, blue = PP, purple = SQ).


Some separation of values relative to population origin of *D. mojavensis* is evident, with PC 1 and PC 2 displaying the most significant grouping. The visualization (Fig. 2) further depicts that CHC composition loads heavily on PC1 and PC 2, which we can also conclude from the scree plot (Fig. 1). This indicates that reaction to desiccation stress measured by change in CHC composition can be differentiated between *D. mojavensis* most effectively on the basis of population origin.


Comparing the bivariate plots relative to **host plant**: 

```{r}
c1_2 <- ggplot(aes(x=Comp.2,y=Comp.1, colour=Cactus), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 2") + theme(legend.position = "none")

c1_3 <- ggplot(aes(x=Comp.3,y=Comp.1, colour=Cactus), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 3") + theme(legend.position = "none")

c1_4 <- ggplot(aes(x=Comp.4,y=Comp.1, colour=Cactus), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 4") + theme(legend.position = "none")

c2_3 <- ggplot(aes(x=Comp.3,y=Comp.2, colour=Cactus), data=CHCDat) + geom_point() + ylab("PC 2") + xlab("PC 3") + theme(legend.position = "none")

c2_4 <- ggplot(aes(x=Comp.4,y=Comp.2, colour=Cactus), data=CHCDat) + geom_point() + ylab("PC 2") + xlab("PC 4") + theme(legend.position = "none")

c3_4 <- ggplot(aes(x=Comp.4,y=Comp.3, colour=Cactus), data=CHCDat) + geom_point() + ylab("PC 3") + xlab("PC 4") + theme(legend.position = "none")

PCPlot_Cac <- grid.arrange(c1_2, c1_3, c1_4, c2_3, c2_4, c3_4, nrow=2)

```

**Fig. 3.** All possible combinations of principle components 1-4 with points coloured by host plant (red = AG, blue = OP).


No significant separation of values is observed relative to host plant. 


Comparing the bivariate plots relative to desiccation **exposure time**: 

```{r}
e1_2 <- ggplot(aes(x=Comp.2,y=Comp.1, colour=Exposure), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 2") + theme(legend.position = "none")

e1_3 <- ggplot(aes(x=Comp.3,y=Comp.1, colour=Exposure), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 3") + theme(legend.position = "none")

e1_4 <- ggplot(aes(x=Comp.4,y=Comp.1, colour=Exposure), data=CHCDat) + geom_point() + ylab("PC 1") + xlab("PC 4") + theme(legend.position = "none")

e2_3 <- ggplot(aes(x=Comp.3,y=Comp.2, colour=Exposure), data=CHCDat) + geom_point() + ylab("PC 2") + xlab("PC 3") + theme(legend.position = "none")

e2_4 <- ggplot(aes(x=Comp.4,y=Comp.2, colour=Exposure), data=CHCDat) + geom_point() + ylab("PC 2") + xlab("PC 4") + theme(legend.position = "none")

e3_4 <- ggplot(aes(x=Comp.4,y=Comp.3, colour=Exposure), data=CHCDat) + geom_point() + ylab("PC 3") + xlab("PC 4") + theme(legend.position = "none")

PCPlot_Exp <- grid.arrange(e1_2, e1_3, e1_4, e2_3, e2_4, e3_4, nrow=2)

```

**Fig. 4.** All possible combinations of principle components 1-4 with points coloured by exposure time length in hours (red = 0 h, green = 9 h, blue = 18 h). 


No significant separation of values is observed relative to desiccation exposure time.

## Part 2: Decision tree

To determine which cuticular hydrocarbon changes predict dessication, we will create a decision tree. This will allow us to determine which CHCs are involved in the dessication response in _D. mojavensis_.

## Setup

Loading libraries:
```{r}
library(dplyr)
library(pROC)
library(e1071)
library(tree)
library(rpart)
library(gbm)
library(randomForest)
library(ggplot2)
library(tibble)
library(ggpubr)
```

Loading theme:
```{r}
library(cowplot)
theme_set(theme_cowplot(12))
```


## Loading the data
```{r}
CHC_dat<-read.csv("./data/CHC_dat.csv")
```

## Inspecting the data
```{r}
names(CHC_dat)
```

There are a lot of extra columns in the original data file that need to be removed (everything after "TOTAL")

```{r}
CHC_dat <- CHC_dat[1:37]
names(CHC_dat)
```
```{r}
str(CHC_dat)
```

Sample, pop, sex, cactus, and exposure should all be set to factors, and C.37.br.alkene should be set to numeric:
```{r}
CHC_dat <- CHC_dat %>% 
  mutate(SAMPLE = as.factor(SAMPLE),
         pop = as.factor(pop),
         sex = as.factor(sex),
         cactus = as.factor(cactus),
         exposure = as.factor(exposure),
         C.37.br.alkene = as.numeric(C.37.br.alkene)) #edit by Sam; CHC content changed from character to numeric
```

Checking the updated structure:
```{r}
str(CHC_dat) #everything here looks good
```

```{r}
head(CHC_dat)
```

```{r}
tail(CHC_dat)
```

Seeing if there are NAs:

```{r}
# Seeing which columns have missing values
CHC_dat %>%
  select_if(function(x) any(is.na(x))) %>%
  names() 
```

There was a single missing value, represented by a period, that was coerced into an NA when the column was converted from character to numeric.

Removing row containing NA value:

```{r}
CHC_dat<-CHC_dat[complete.cases(CHC_dat),]
```

## Dimensions of the data frame
```{r}
dim(CHC_dat)
```

The final dimensions are 115 rows and 37 columns.


## Splitting the data into a training and a test dataset
```{r}
Rows<-c(1:nrow(CHC_dat))
Train<-Rows %% 2==1
Validate<-Rows %% 2==0

dat_train<-CHC_dat[Train,]
dat_validate<-CHC_dat[Validate,]

head(dat_train)
head(dat_validate)
```


## Decision tree

### Tree diagram with text

Getting only response and predictor variables:
```{r}
names(dat_train)
```

* This has Sample, pop, sex, cactus, and exposure, we want only exposure

```{r}
# Selecting only response and predictor variables:
dat_train_r <- dplyr::select(dat_train, -c("SAMPLE":"cactus"))
head(dat_train_r)
```

Creating the tree:
```{r}
DryTree<-tree(exposure~., data=dat_train_r)
plot(DryTree)
text(DryTree, cex=0.7, adj=0)
```

> A decision tree of the CHC features involved in response to different levels of dessication. Outcomes are printed along the bottom of the tree, coming from 'roots' and nodes listing each CHC feature & its corresponding value for decision-making.


### CHC feature most influential for classifying samples
The protein feature that was most influential for classifying samples was C.34. (at the top of the tree).

### Using predict() to provide the confusion matrix
```{r}
# Selecting only response and predictor variables:
dat_validate_r <- dplyr::select(dat_validate, -c("SAMPLE":"cactus"))

CatDat<-data.frame(Obs=dat_validate_r$exposure, Pred=predict(DryTree, dat_validate_r, type="class"))
table(CatDat)
```

## 4. Finding the misclassification error rate
```{r}
MisClass<-CatDat %>%
  filter(Obs!=Pred)
```


```{r}
nrow(MisClass)/nrow(CatDat)
```

The misclassification error rate was ~60%. This is a high error rate.

## Random Forest

Running randomforest function on training dataset, excluding rows with NA values:

```{r}
noNA<-complete.cases(dat_train_r)
dat_train_final<-dat_train_r[noNA,]
DryFor<-randomForest(exposure ~ ., data=dat_train_final, ntree=100, mtry=3, nodesize=5, importance=TRUE)
DryFor
```
Getting classifications Using predict():

```{r}
PredFor<-predict(DryFor, dat_validate_r, type="class")
head(DryFor)
```

Getting confusion matrix:
```{r}
CatDat2<-data.frame(Obs=dat_validate_r$exposure, Pred=predict(DryFor, dat_validate_r, type="class"))
table(CatDat2)
```

Getting the misclassification error rate:
```{r}
MisClass2<-CatDat2 %>%
  filter(Obs!=Pred)
nrow(MisClass2)/nrow(CatDat2)
```

The misclassification rate has dropped from ~60% to ~42% by using random forests.

### Plotting
```{r}

dry_imp<-as.data.frame(DryFor$importance)
dry_imp<-rownames_to_column(dry_imp, "CHC")
dry_imp<-dry_imp


d_imp <- ggplot(data=dry_imp, aes(x=reorder(CHC, MeanDecreaseAccuracy), y=MeanDecreaseAccuracy)) +
  geom_col() +
  xlab("CHC") +
  ylab("Importance")
ggpar(d_imp, x.text.angle=45)
```

> The relative importance of different CHC features in response to different levels of dessication

# Question 2 Analysis: Investigating the influences of sex, population, and cactus species on the viability of _Drosophila mojavensis_ eggs.

For this question, we set out to determine whether **sex**, **population origin**, or **host cactus species** affected the viability of _D. mojavensis_ eggs. To do so, we will first construct a linear mixed-effects model of the above factors and use AIC criteria to find the most optimal one. We will also use a decision tree as a second method of analyzing these factors. From this, we will determine what influences the viability of _D. mojavensis_, which may have important implications for predicting the capacity of the species to survive under changing environments.


## Loading libraries & data

Libraries:
```{r}
library(ggplot2)
library(dplyr)
library(lme4)
library(cowplot)
theme_set(theme_cowplot(12))
library(MuMIn)
library(MASS)
```

Data:
```{r}
via_dat <- read.csv("./data/Viability_data.csv")
```

## Checking the structure of the data
```{r}
head(via_dat)
```

```{r}
tail(via_dat)
```

```{r}
str(via_dat)
```

```{r}
names(via_dat)
```


## Transformations for quality of data
```{r}
# Removing the blank rows at the end of the dataset
via_dat <- slice(via_dat, 1:24)
tail(via_dat)
```

```{r}
# Renaming the columns for better formatting
via_dat <- rename(via_dat, total_male = total.Male,
       total_female = total.Female,
       total_flies = TOTAL,
       total_hatching = Egg.Hatching,
       viability = Viability....)
names(via_dat)
```

```{r}
# Structuring the columns correctly
via_dat <- via_dat %>% mutate(
  Bottle = as.factor(Bottle),
  Population = as.factor(Population),
  Cactus = as.factor(Cactus)
)

str(via_dat)
```

## Checking the viability data for assumptions of normality
```{r}
ggplot(via_dat, aes(x=viability)) +
  geom_histogram() 
```

Although not perfect, the data does fit within a mostly normal distribution. So, we don't need to use a generalized linear model with any other distribution parameters.

Next, we want to have sex as a predictor, but this is currently split into two separate categories (total_male and total_female). To include it, this should be restructured as a single column. For this, we'll choose to make a column called ``sex_bias`` measuring the deviation of the proportion of male and female flies from 50:50. If, for instance, viability is higher when the proportion is skewed towards male flies, then that may indicate that viability is influenced by sex.

```{r}
via_dat <- via_dat %>% mutate(
  sex_bias = 0.5 - (total_male/total_flies)
) # Negative value indicates female bias
head(via_dat)
```


## Constructing the linear model

```{r}
tail(via_dat)
full_model <- lm(viability ~ Cactus + sex_bias + Population, data=via_dat) # Excluding bottle because each observation has a unique bottle number, returns an error otherwise
```


## Model selection

For this, we will use backwards selection to determine the best-fit model from all possible predictors.

```{r}
BackSel <- stepAIC(full_model, direction="backward")
```


According to the AIC criteria, there are no significant differences between models (they're all within ~2 of each other).


## ANOVA for significance of each factor
```{r}
anova(full_model)
```

In agreement with the AIC, the only factor that has a significant effect on egg viability is the population of origin.

## QA/QC
```{r}
qplot(x=residuals(full_model))
```

The residuals are fairly normally distributed, meeting assumptions of normality. 


