---
title: "Dryfly_Q1"
author: "Isabeau Lewis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1 Analysis: Investigating the influences of sex, population, and cactus species on the viability of _Drosophila mojavensis_ eggs.

For our first question, we set out to determine whether **sex**, **population origin**, or **host cactus species** affected the viability of _D. mojavensis_ eggs. To do so, we will first construct a linear mixed-effects model of the above factors and use AIC criteria to find the most optimal one. We will also use a decision tree as a second method of analyzing these factors. From this, we will determine what influences the viability of _D. mojavensis_, which may have important implications for predicting the capacity of the species to survive under changing environments.


## Loading libraries & data

Libraries:
```{r}
library(ggplot2)
library(dplyr)
library(lme4)
library(cowplot)
theme_set(theme_cowplot(12))
library(MuMIn)
library(MASS)
```

Data:
```{r}
via_dat <- read.csv("./data/Viability_data.csv")
```

## Checking the structure of the data
```{r}
head(via_dat)
```

```{r}
tail(via_dat)
```

```{r}
str(via_dat)
```

```{r}
names(via_dat)
```


## Transformations for quality of data
```{r}
# Removing the blank rows at the end of the dataset
via_dat <- slice(via_dat, 1:24)
tail(via_dat)
```

```{r}
# Renaming the columns for better formatting
via_dat <- rename(via_dat, total_male = total.Male,
       total_female = total.Female,
       total_flies = TOTAL,
       total_hatching = Egg.Hatching,
       viability = Viability....)
names(via_dat)
```

```{r}
# Structuring the columns correctly
via_dat <- via_dat %>% mutate(
  Bottle = as.factor(Bottle),
  Population = as.factor(Population),
  Cactus = as.factor(Cactus)
)

str(via_dat)
```

## Checking the viability data for assumptions of normality
```{r}
ggplot(via_dat, aes(x=viability)) +
  geom_histogram() 
```

Although not perfect, the data does fit within a mostly normal distribution. So, we don't need to use a generalized linear model with any other distribution parameters.

Next, we want to have sex as a predictor, but this is currently split into two separate categories (total_male and total_female). To include it, this should be restructured as a single column. For this, we'll choose to make a column called ``sex_bias`` measuring the deviation of the proportion of male and female flies from 50:50. If, for instance, viability is higher when the proportion is skewed towards male flies, then that may indicate that viability is influenced by sex.

```{r}
via_dat <- via_dat %>% mutate(
  sex_bias = 0.5 - (total_male/total_flies)
) # Negative value indicates female bias
head(via_dat)
```


## Constructing the linear model

```{r}
tail(via_dat)
full_model <- lm(viability ~ Cactus + sex_bias + Population, data=via_dat) # Excluding bottle because each observation has a unique bottle number, returns an error otherwise
```


## Model selection

For this, we will use backwards selection to determine the best-fit model from all possible predictors.

```{r}
BackSel <- stepAIC(full_model, direction="backward")
```


According to the AIC criteria, there are no significant differences between models (they're all within ~2 of each other).


## ANOVA for significance of each factor
```{r}
anova(full_model)
```

In agreement with the AIC, the only factor that has a significant effect on egg viability is the population of origin.

## QA/QC
```{r}
qplot(x=residuals(full_model))
```

The residuals are fairly normally distributed, meeting assumptions of normality. 
